Toward theories of deliberated preferences

### 2020-10-17 simple.tex
Commentaires de Yves reçus par e-mail.

\section{Motivation} 
It has escaped nobody’s attention that 
studying reflective preferences, in supplement to shallow preferences, is an important endeavor for economics: we want to recommend to individual decision makers; we want to adopt optimization criteria for society, and this ought not to merely faithfully reproduce unreflectful behavior. This is one important difference between the normative and the (inappropriately called, as will be seen) “descriptive” approach in economics. As \citet{hausman_preference_1994} illustrate with an example of a person drinking a fatal poison while erroneously thinking that it is water, shallow preferences are not sufficient to inform normative economics.
Yet, classical methods to obtain preference models in economics are either formally defined on the basis of observable data that do not describe reflective preferences, or are obtained by other means than observable preferential data and systematic reproducible formal procedures.

This article is interested in defining theories of decision making based on another kind of observable data, namely, the reactions of decision makers to arguments. This permits to overcome a current important limitation of normative approaches, that of being not empirical (in a sense that will be developed below).

By reflective preferences (of an individual), I mean here preferences that represent informed, pondered decisions, that would be taken after due consideration by the individual. Reflective preferences can be contrasted to shallow preferences, those that \citet[p.\ 16]{von_neumann_theory_2004} describe as “intuitive”, as originating from an “immediate sensation” (cited in \citet{fishburn_retrospective_1989}).
This distinction draws on our capacity, as human beings, to think about our shallow preferences and consider some of them as mistaken \citep{frankfurt_freedom_1971}. A famous example of such a situation is the one of \citet[pp.\ 101--103]{savage_foundations_1972} having realized, after due consideration, that his initial declared preference in a decision problem presented by \citet{allais_so-called_1979} was mistaken.

It is well known that shallow preferences do not always match reflective ones.
Indeed, forming reflective preferences is recognised as a difficult problem in many decision problems, contrary, by definition, to shallow preferences. For example, \citet[p.\ 6]{colell_microeconomic_1995} write, when presenting the preference relation that is the basis for most of their famous book about Microeconomic theory: “Introspection quickly reveals how hard it is to evaluate alternatives that are far from the realm of common experience. It takes work and serious reflection to find out one’s own preferences.” 
In the words of \citet[p.\ 9]{bell_descriptive_1988}, “Our decision makers are not economic automatons; they make mistakes, have remorse, suffer anxieties, and cannot make up their minds” (cited by \citet[p.\ 87]{deparis_etude_2012}).
Indeed, it is to be expected that some behavior highlighted by psychologists \citep{kahneman_thinking_2012}, such as framing effects, would be rejected by the individuals themselves when pondering their decisions.

That reflective preferences are relevant to economic theory, in supplement to shallow preferences, is also illustrated by the general requirement of transitivity of preferences for economic theory, which is reasonably expected from reflective rather than from shallow preferences. As an illustration, \citet[p.\ 7]{colell_microeconomic_1995} go on writing: “the transitivity assumption can be hard to satisfy when evaluating alternatives far from common experience. As compared to the completeness property, however, it is also more fundamental in the sense that substantial portions of economic theory would not survive if economic agents could not be assumed to have transitive preferences.”

@@

\subsection{Existing approaches for capturing reflective preferences}
\subsubsection{Laundering preferences}
An approach much discussed recently is the one of laundering preferences, also called preference purification \citep{bernheim_beyond_2009, rubinstein_eliciting_2011, infante_preference_2016}: having collected shallow preference data (from the field or from interviews), laundering preferences consists in transforming them, using a formal process, in such a way that they satisfy axioms generally considered as required by rationality, or keeping only preference data collected in some situations considered appropriate to determine welfare-relevant preferences. For example, \citet{bleichrodt_making_2001} transform observed preferences in order to obtain a classical Expected Utility model; \citet{bernheim_beyond_2009} propose to exclude situations where the individual seems (by objective criteria) unable to process information considered (by the theorist) as relevant information.

This creates a risk of mis-representing the reflective preferences of the individual: not only is it not known whether the individual would reflectively want to adhere to these axioms and, if so, how she would reflectively adapt her preferences when her shallow preferences exhibit related inconsistencies; but it is not even discussed how in principle this could be known. As a result, a theory postulating constraints on individual’s behavior on the basis of rational requirements is subject to the criticism addressed by \citet[p.\ 228]{dold_toward_2018} to libertarian paternalism, in that it “effectively denies the possibility that deviations from rational choice might be something to be acknowledged as reasonable”. 
That this is an important methodological problem is sometimes recognized by authors adopting this approach. \Citet{bleichrodt_making_2001}, for example, state that laundering require controversial assumptions (p.\ 1500), and recommend it only as a last resort, recommending discussing with the decision-maker whenever possible to elicit her reflective preferences (p.\ 1499). Note that nudging \citep{thaler_nudge_2009} creates a similar problem: we have to guess what the individual would like if he thought hard about the decision problem, was well-informed, and actively worked against the psychological features he himself does not consider adequate for good decision making.

Some axioms arguably seem innocuous. For example, Pareto-dominance requires that $a$ be weakly preferred to $b$ if $a$ is at least as good as $b$ on every point of view that matter. However, as has been amply remarked, what is required is not merely the correctness of the axiom in the abstract, but its correct application to the decision situation. Has a point of view that the decision maker considers relevant been forgotten, leading the analyst to mistakenly think that $a$ Pareto-dominates $b$, and the seemingly impeccable conclusion that $a$ ought to be preferred to $b$ turns incorrect. \citet{sen_maximisation_1997} illustrated this reasoning (with another axiom) with his well-known example of a person picking her second most-preferred fruit from a basket out of politeness reasons, which could be considered irrational by an analyst blindly applying WARP. As \citet[p.\ 40]{lecouteux_reconciling_2015} puts it (citing \citet[p.\ 13]{bacharach_beyond_2006} for the part in italics), and as analyzed by \citet{sen_information_1986, sen_internal_1993}, “whether a decision-theoretic principle of rationality has been violated \emph{depends on how we, the theorists, ‘cut up the world’}” \TODO{Find that book}.

\subsubsection{Informal deliberation}
Another path to obtain reflectful preferences has been followed: the one of deliberation, an umbrella term by which I designate processes, either, involving discussions between an analyst and a decision-maker while querying her preferences in view of building a formal model of it \citep{raiffa_back_1985, keeney_decisions_1993, roy_multicriteria_1996, belton_multiple_2002}, or involving group discussions, with presentations from experts to an audience and internal deliberation phases \citep{fishkin_when_2011}, aiming to obtain well-informed preferences.

A major problem of these techniques is that they are not easily reproducible, and that the resulting preferences elicited from the subjects might reflect properties of the personalities of the intervening experts or group-dynamic effects rather than true reflective preferences of the individuals \commentOCf{Yves, as-tu une référence pour moi ?}. This problem is exacerbated by the fact that psychologists have shown that apparent coherence can appear, that is, preferences that do not violate axioms usually considered as representing rationality, although some framing effect may be at play that result in the apparently coherent preferences to arguably not represent reflective preferences \citep{ariely_predictably_2010}. It also remains to be clarified how to check on a principled, systematic basis, whether all relevant arguments have been presented faithfully. Comfronted with (possibly self-designated) experts with enough assertiveness, a decision-maker can be convinced to follow principles of decision-making that several experts in the field of decision-theory consider incorrect. As an illustration, after discussing the weaknesses of the \ac{AHP}, \citet[p.\ 53]{howard_foundations_2007} write: “Why, then, do inferior processes find favor with decision makers? The answer is that they do not force you to think very hard or to think in new ways.”
Beyond the controversial \ac{AHP}, famous decision analysts have voiced disagreements with normative theories of decision making \citep{ellsberg_risk_1961, allais_so-called_1979} that are however generally applied in the field. 
\Citet[p.\ 669]{ellsberg_risk_1961} concluded his influential article as follows: “[In some situations, for some people,] the Bayesian or Savage approach gives [...] bad advice. [Some people] act in conflict with the axioms deliberately, without apology, because it seems to them the sensible way to behave. Are they clearly mistaken?” 
This rethorical question illustrates the need of systematic procedures able to determine whether decision-makers, when confronted with the relevant arguments, accept the soundness of the principles on whose basis the recommendation rests (this has been somewhat explored, but not systematically, for current lack of a systematic way of confronting decision makers to arguments \citep{slovic_who_1974, stanovich_discrepancies_1999}).

As a consequence of these techniques being not completely formalised and reproducible, each application requires manual intervention of (very skilled) experts. Eliciting reflective preferences therefore requires enormous resources. \Citet[p.\ 572]{pinto_normative_2012} observe that “in many of those surveys that governments use in order to inform social policy there are not enough resources (time and money) in order to reach the stage where preferences are not labile”. There is, to reduce this cost, a need for theories that can be iteratively improved upon and yield reusable knowledge about reflective preferences.

Different analysts, using different methods and different axioms as their favorite conception of rational decision-making, will obtain possibly contradictory conclusions, “elicited” from the decision-makers who hire them. Some analysts admit this possibility and do not consider it to be a problem.
For example, \citet[p.\ 214]{roy_comparison_1995} write: “In our view, these inevitable disagreements do not imply that decision-aid is useless but simply that a single problem may have several valid responses. Given that two different decision-aid models cannot be implemented in the same decision process, the decision-maker must be conscious of the qualitative choices implied by the different models – often conveying the analysts’ own ethical choices – before coming to personal conclusions on the choice to be made. In this domain, the many different approaches reflect in our view the complexity of the researcher's task much more than a scientific weakness.”
Although being an important insight about the (likely vain) hope of determining a single correct recommendation in many complex decision situations (about which more below), this answer does not address the problem that some principles of decision making may be wrong guides in some situations, possibly including some principles that are commonly applied in the field, hence the need of principled ways of discriminating them.

\subsection{Some other related approaches}
The prescriptive approach to decision making \citep{bell_decision_1988} has been proposed (in supplement to the normative one and the descriptive one), which insists on incorporating the findings of psychologists about decision maker’s weaknesses of reasoning in the procedures leading to recommendations. I do not discuss this approach specifically here, because its methods can be considered to fall into the two categories I have described here above when it aims at obtaining reflective preference models. When considering normative economics broadly, as I do here, as the methods aiming for valid prescription for an individual or a group, prescriptive approaches are a part of normative economics.

Some authors have proposed to ground normative theories on observable aspects of nature and individuals, to solve some of the problems mentioned above; for example, \citet{kahneman_back_1997} propose to measure “experienced utility”, considered as a hedonic quality. To the best of my knowledge, these proposals never aim at observing reflective preferences as considered here and thus fall outside the scope of this article.

The field of formal argumentation theory has developed technical approaches to deal with arguments. Such endeavor starts with an “attack” relation, which indicates which arguments argue against which. Various criteria can then be mathematically defined, in the form of a function selecting subsets of arguments given an attack relation, typically representing sets of arguments that “best resist” their counter-arguments. This field has greatly inspired this proposal, in particular, the notion of attack and of reinstatement. This field however does not view the attack relation as observable, but considers it as given, and its methodology is therefore not empirical in the sense seeked for here. (\Citet{cailloux_formal_2019} elaborates on this difference of objective.)

An important literature exists about “debiasing techniques”: techniques that aim at encouraging individuals to think more about their decision problem in order to reduce occurrence of behavioral “mistakes”. Examples include suggesting the individuals to consider alternative points of view, or asking them to give reasons for their choices (an insightful article by \citet{mitchell_libertarian_2004} cite many such studies). This proposal is situated at another level, which could be considered as a way of justifying and comparing such debiasing techniques with no need of a prior agreement on what constitutes a behavioral mistake.

The current proposal pursues a path started with \citet{cailloux_formal_2019}, an article that defines the notion of \ac{DJ} and models that enable interactions with individuals whose \acp{DJ} are to be captured. That article however requires as fundamental observable a “trump” relation, indicating whether an individual thinks that an argument is against another argument; and a “support” relation, indicating whether an individual thinks that an argument supports a proposition. The current proposal defines every required concepts, such as the notion of arguments attacking other arguments, on the basis of a single observable that is compatible with the revealed preference approach priviledged in economic theory, in the sense that it can be defined by reference to what individuals effectively choose under various circumstances. The current proposal also defines explicitly (in the form of an axiom) sufficient requirements to grant normative power to observations, an aspect which was previously left unclear. Finally, the current proposal links \ac{DJ} and deliberated preferences.

@@

With no intent of reducing the importance of these findings, it is important to qualify such statements. In supplement to them being not necessarily applicable to reflective preferences, the findings of psychologists do not, evidently, show that all preference statements are labile. As \citet[p.\ xvi]{lichtenstein_construction_2006} write, “To be sure, there are limits to the process of construction. Some televisions just won’t sell; most of the time, people are unlikely to want 80\% of their wages to go into savings. It would be an overstatement to say of preferences, as Gertrude Stein said of Oakland, that “there is no there there.”

>>>>

\section{Motivation} 
\commentYM{dans la mesure où tu dis concevoir ou avoir conçu ce draft comme étant presque soumettable, pourquoi as-tu cette section "motivation" plutôt qu'une classique introduction ?}}
It has escaped nobody’s attention that 
studying reflective preferences, in supplement to shallow preferences, is an important endeavor for economics: we want to recommend to individual decision makers; we want to adopt optimization criteria for society, and this ought not to merely faithfully reproduce unreflectful behavior.\commentYM{je trouve cette entame provocatrice assez maladroite, car elle s'appuie sur l'opposition reflective vs. shallow preference, qui ne parle (je pense) pas immédiatement au lecteur. Je pense indispensable de définir cette dichotomie avant d'en parler. Par ailleurs que veut dire "is an important endeavor for economics" ? Il me semble au contraire que c'est une tâche qui devrait tenir à coeur aux économistes, mais qui pourtant n'est pas pour eux une tâche à laquelle ils s'attèlent vigoureusements. Enfin, l'anglais de cette phrase me semble un peu maladroit} This is one important difference between the normative and the (inappropriately called, as will be seen) “descriptive” approach in economics\commentYM{où discutes-tu cette distinction, dans son application à l'économie en particulier ?}. As \citet{hausman_preference_1994} illustrate with an example of a person drinking a fatal poison while erroneously thinking that it is water, shallow preferences are not sufficient to inform normative economics.
Yet, classical methods to obtain preference models in economics are either formally defined on the basis of observable data that do not describe reflective preferences, or are obtained by other means than observable preferential data and systematic reproducible formal procedures. \commentYM{je devine ce que tu veux dire, mais je ne comprends pas la dernière phrase. Il me semble important que tu t'expliques ici, car je pense que le lecteur ici ne peut pas comprendre la niche dans laquelle s'inscrit ton projet}

This article is interested in defining theories of decision making based on another kind of observable data\commentYM{other than what? Juste avant tu dis que les autres s'appuient sur des données qui ne reflètent pas les préférences qui t'intéressent, et maintenant tu dis que tu vas regarder autre chose qu'eux. Cela donne l'impression que tu tournes autour du pot, car tu ne dis pas ce que les autres regardes et pourquoi ça ne t'intéresse pas}, namely, the reactions of decision makers to arguments. This permits to overcome a current important limitation of normative approaches, that of being not empirical (in a sense that will be developed below).\commentYM{phrase incompréhensible pour le lecteur, puisque tu dis toi même que tu en expliqueras les termes plus tard. C'est une rhétorique difficile à tenir dans une introduction}

By reflective preferences (of an individual), I mean here preferences that represent informed, pondered decisions, that would be taken after due consideration by the individual. Reflective preferences can be contrasted to shallow preferences, those that \citet[p.\ 16]{von_neumann_theory_2004} describe as “intuitive”, as originating from an “immediate sensation” (cited in \citet{fishburn_retrospective_1989}).
This distinction draws on our capacity, as human beings, to think about our shallow preferences and consider some of them as mistaken \citep{frankfurt_freedom_1971}. \commentYM{ces définitions devraient être mises + haut, car le lecteur ne peut pas comprendre ce que tu racontes sans elles}A famous example of such a situation is the one of \citet[pp.\ 101--103]{savage_foundations_1972} having realized, after due consideration, that his initial declared preference in a decision problem presented by \citet{allais_so-called_1979} was mistaken.

It is well known that shallow preferences do not always match reflective ones.\commentYM{à quoi sert la phrase précédente ?}
Indeed, forming reflective preferences is recognised as a difficult problem in many decision problems, contrary, by definition, to shallow preferences. For example, \citet[p.\ 6]{colell_microeconomic_1995} write, when presenting the preference relation that is the basis for most of their famous book about Microeconomic theory: “Introspection quickly reveals how hard it is to evaluate alternatives that are far from the realm of common experience. It takes work and serious reflection to find out one’s own preferences.” 
In the words of \citet[p.\ 9]{bell_descriptive_1988}, “Our decision makers are not economic automatons; they make mistakes, have remorse, suffer anxieties, and cannot make up their minds”\commentYM{cette citation me semble mal venue ici, car elle ne pose pas que la question du contraste shallow/reflective, mais aussi la question de savoir s'il faudrait toujours faire un calcul économique. Par exemple avoir des remords ne me semble pas forcément une mauvaise chose} (cited by \citet[p.\ 87]{deparis_etude_2012}).
Indeed, it is to be expected that some behavior highlighted by psychologists \citep{kahneman_thinking_2012}, such as framing effects, would be rejected by the individuals themselves when pondering their decisions. \commentYM{formulation ambigue : on dirait que tu dis que les gens nieraient que les framing effects existent}

That reflective preferences are relevant to economic theory, in supplement to shallow preferences, is also illustrated by the general requirement of transitivity of preferences for economic theory, which is reasonably expected from reflective rather than from shallow preferences. As an illustration, \citet[p.\ 7]{colell_microeconomic_1995} go on writing: “the transitivity assumption can be hard to satisfy when evaluating alternatives far from common experience. As compared to the completeness property, however, it is also more fundamental in the sense that substantial portions of economic theory would not survive if economic agents could not be assumed to have transitive preferences.” \commentYM{la seconde phrase de la citation me parait superflue pour ce que tu as à dire ici. Plus généralement, la plus-value d'une citation ici me semble limitée. Plus fondamentalement, l'argument me semble incomplet. OK, les préférences shallow ont toutes les chances de n'être pas transitives. Mais pourquoi penser que les préférences réfléchies le seraient ? Il manque un argument}

@@

\subsection{Existing approaches for capturing reflective preferences}
\subsubsection{Laundering preferences}
An approach much discussed recently is the one of laundering preferences, also called preference purification \citep{bernheim_beyond_2009, rubinstein_eliciting_2011, infante_preference_2016}: having collected shallow preference data (from the field or from interviews), laundering preferences consists in transforming them, using a formal process, in such a way that they satisfy axioms generally considered as required by rationality, or keeping only preference data collected in some situations considered appropriate to determine welfare-relevant preferences. For example, \citet{bleichrodt_making_2001} transform observed preferences in order to obtain a classical Expected Utility model; \citet{bernheim_beyond_2009} propose to exclude situations where the individual seems (by objective criteria) unable to process information considered (by the theorist) as relevant information. \commentYM{les deux mécanismes évoqués sont très différents, et ont des enjeux différents. Je pense qu'il serait utile de les distinguer, et je pense qu'il faudrait en dire plus sur la nature de la transformation et du filtrage, sans quoi la discussion des ces approches est trop superficielle}

This \commentYM{what? ici on voit qu'on gagnerait à savoir concrètement quelle est la manip déployée par les auteurs ci-dessus}creates a risk of mis-representing the reflective preferences of the individual: not only is it not known whether the individual would reflectively want to adhere to these axioms \commentYM{quels axiomes ? besoin d'expliquer + en détail ce que font ces auteurs ; et tu sembles pressupposer ici qu'adéhrer aux axiomes et s'approprier une transformation ou un filtre basé sur des axiomes, c'est la même chose, ce qui ne me semble pas évident}and, if so, how she would reflectively adapt her preferences when her shallow preferences exhibit related inconsistencies; but it is not even discussed how in principle this could be known \commentYM{phrase en + super longue : arrivée à la fin, je suis perdu}. As a result, a theory postulating constraints on individual’s behavior \commentYM{je ne comprends pas de quelles contraintes sur le comportement il est question ici. Les théories des laundered preferences que je connais n'imposent aucune contrainte au COMPORTEMENT. Cela souligne encore une fois qu'il faudrait être plus prolixe sur la description des approches dont tu parles}on the basis of rational requirements is subject to the criticism addressed by \citet[p.\ 228]{dold_toward_2018} to libertarian paternalism, in that it “effectively denies the possibility that deviations from rational choice might be something to be acknowledged as reasonable”. \commentYM{phrase bien alambiquée, qui fait référence à un concept ("libertarian paternalism") que je pense abscond pour bien des gens (dont moi). Je pense que tu devrais reformuler, donner l'argument plus directement, et si tu y tiens mettre entre parenthèse que cet arguement est utilisé dans un autre débat par x et y quand ils parlent de LP}
That this is an important methodological problem is sometimes recognized by authors adopting \commentYM{endorsing}this approach \commentYM{je pense qu'en anglais il est + idiomatique d'éviter ce genre de tournure : plutôt dire "Authors... sometimes recognize that..."}. \Citet{bleichrodt_making_2001}, for example, state that laundering require controversial assumptions (p.\ 1500), and recommend it only as a last resort, recommending discussing with the decision-maker whenever possible to elicit her reflective preferences (p.\ 1499). Note that nudging \citep{thaler_nudge_2009} creates a similar problem: we have to guess what the individual would like if he thought hard about the decision problem, was well-informed, and actively worked against the psychological features he himself does not consider adequate for good decision making.

Some axioms arguably seem innocuous. For example, Pareto-dominance requires that $a$ be weakly preferred to $b$ if $a$ is at least as good as $b$ on every point of view that matter. However, as has been amply remarked, what is required is not merely the correctness of the axiom in the abstract, but its correct application to the decision situation. Has a point of view that the decision maker considers relevant been forgotten, leading the analyst to mistakenly think that $a$ Pareto-dominates $b$, and the seemingly impeccable conclusion that $a$ ought to be preferred to $b$ turns incorrect. \citet{sen_maximisation_1997} illustrated this reasoning (with another axiom) with his well-known example of a person picking her second most-preferred fruit from a basket out of politeness reasons, which could be considered irrational by an analyst blindly applying WARP. As \citet[p.\ 40]{lecouteux_reconciling_2015} puts it (citing \citet[p.\ 13]{bacharach_beyond_2006} for the part in italics), and as analyzed by \citet{sen_information_1986, sen_internal_1993}, “whether a decision-theoretic principle of rationality has been violated \emph{depends on how we, the theorists, ‘cut up the world’}” \TODO{Find that book}.\commentYM{Je § tout entier me semble mal placé ici. Les considérations évoquées sont relatives au choix d'un "bon" axiome, problématique classique dans les discussions des théories du choix rational, mais dont le lien avec le laundering des préférences ne crève pas les yeux. Il y a des moyens de faire le lien, mais là tu n'expliques aucun lien. Donc ton § tombe comme un cheveux sur la soupe et le lecteur est embrouillé, ne se rappelle même plus quel était l'object de la sous-sous section qu'il vient de lire}

\subsubsection{Informal deliberation}
Another path to obtain reflectful \commentYM{reflective ?} preferences has been followed: the one of deliberation, an umbrella term by which I designate processes, either, involving discussions between an analyst and a decision-maker while querying her preferences in view of building a formal model of it \commentYM{them ?} \citep{raiffa_back_1985, keeney_decisions_1993, roy_multicriteria_1996, belton_multiple_2002}, or involving group discussions, with presentations from experts to an audience and internal deliberation phases \citep{fishkin_when_2011}, aiming to obtain well-informed preferences. \commentYM{il me semble nécessaire ici de mentionner aussi tous les dispositifs participatifs et de concertation employés en routine dans les prises de décisions relatives aux politiques publiques et leur mise en oeuvre}

A major problem of these techniques is that they are not easily reproducible \commentYM{il faudrait expliquer en quel sens}, and that the resulting preferences elicited from the subjects might reflect properties of the personalities of the intervening experts or group-dynamic effects rather than true reflective preferences of the individuals \commentOCf{Yves, as-tu une référence pour moi ?}\commentYM{on a mis des références pertinentes pour les différents aspects dan le papier soumis à Ecol Econ. Ces références ne te conviennent pas ?}. This problem is exacerbated \commentYM{compounded} by the fact that psychologists have shown that apparent coherence can appear, that is, preferences that do not violate axioms usually considered as representing rationality, although some framing effect may be at play that result in the apparently coherent preferences to arguably not represent reflective preferences \citep{ariely_predictably_2010} \commentYM{phrase beaucoup trop longue pour être compréhensible, qui fait par ailleurs référence au concept de rationalité sans le définir, ce qui est acceptable en début d'intro, mais me semble problématique à ce stade dans un papier qui cherche à éclairer les débats flous par des clarifications formelles et conceptuelles}. It also remains to be clarified how to check on a principled, systematic basis, whether all relevant arguments have been presented faithfully. Comfronted with (possibly self-designated \commentYM{ceci est un tout autre problème, dont la mention ici brouille le propos}) experts with enough assertiveness, a decision-maker can be convinced to follow principles of decision-making that several experts in the field of decision-theory \commentYM{would} consider incorrect. As an illustration, after discussing the weaknesses of the \ac{AHP}, \citet[p.\ 53]{howard_foundations_2007} write \commentYM{s}: “Why, then, do inferior processes find favor with decision makers? The answer is that they do not force you to think very hard or to think in new ways.” \commentYM{je ne vois pas en quoi ceci est censé illustrer le propos précédent (ou alors je n'ai rien compris à la citation). J'ai l'impression que la citation dit que, parfois, il y a des forces psychologiques qui s'opposent à l'adoption d'un processus de décision, quand bien même celui-ci aurait des qualités indéniables}
Beyond the controversial \ac{AHP}, famous decision analysts have voiced disagreements with normative theories of decision making \citep{ellsberg_risk_1961, allais_so-called_1979} that are however generally applied in the field. \commentYM{il me semble ici intenable de faire l'économie de définir ce que tu appelles "normative theories"} 
\Citet[p.\ 669]{ellsberg_risk_1961} concluded his influential article as follows: “[In some situations, for some people,] the Bayesian or Savage approach gives [...] bad advice. [Some people] act in conflict with the axioms deliberately, without apology, because it seems to them the sensible way to behave. Are they clearly mistaken?” \commentYM{d'une manière générale, dans toute la section 1, l'avalanche de citation me semble nuire au propos. Cela induit des ruptures dans le style, et force le lecteur à se pencher sur des questions d'interprétation des citations qui l'empêchent de se focaliser sur ton propos à toi}
This rethorical question illustrates the need of systematic procedures able to determine whether decision-makers, when confronted with the relevant arguments, accept the soundness of the principles on whose basis the recommendation rests \commentYM{the principles that the recommendation rests on}(this has been somewhat explored, but not systematically, for current lack \commentYM{for lack} }of a systematic way of confronting decision makers to arguments \citep{slovic_who_1974, stanovich_discrepancies_1999}).

As a consequence of these techniques being not \commentYM{not being}completely formalised and reproducible, each application requires manual intervention \commentYM{quelles sont ces "interventions manuelles" ? à expliquer}of (very skilled) experts. Eliciting reflective preferences therefore requires enormous resources. \Citet[p.\ 572]{pinto_normative_2012} observe that “in many of those surveys that governments use in order to inform social policy there are not enough resources (time and money) in order to reach the stage where preferences are not labile”. There is, to reduce this cost, a need for theories that can be iteratively improved upon and yield reusable knowledge about reflective preferences. \commentYM{Cette idée de limiter le coût est intéressante, et je me dis qu'elle peut faire mouche avec certains lecteurs. Mais en l'état ça ne fonctionne pas car tu ne expliques pas quels sont les coûts impliqués dans les approaches "manuelles"}

Different analysts, using different methods and different axioms as their favorite conception of rational decision-making, will obtain possibly contradictory conclusions, “elicited” from the decision-makers who hire them. Some analysts admit this possibility and do not consider it to be a problem. \commentYM{il me semble que tu soulèves là encore un nouveau problème : comme la notion de rationalité n'est pas claire, il y a débat sur les axiomes qu'on peut vouloir lui associer, et du coup différentes approches, choisissant différents axiomes, vont éliciter des préférences différentes. Je te rejoins sur le fait que c'est un problème important, mais il est différent de ce qui a été discuté juste avant, et tu ne l'explicites pas. Il me semble qu'il faudrait bien distinguer les problèmes, et bien les formuler, et il faudrait que le plan soit plus clair et lisible. Là, en lisant, je ne sais pas si tu considères ce que tu abordes dans ce § comme un nouveau problème, ou comme un aspect ou une implication, du problème exploré précédemment}
For example, \citet[p.\ 214]{roy_comparison_1995} write: “In our view, these inevitable disagreements do not imply that decision-aid is useless but simply that a single problem may have several valid responses. Given that two different decision-aid models cannot be implemented in the same decision process, the decision-maker must be conscious of the qualitative choices implied by the different models – often conveying the analysts’ own ethical choices – before coming to personal conclusions on the choice to be made. In this domain, the many different approaches reflect in our view the complexity of the researcher's task much more than a scientific weakness.”\commentYM{ici encore, je fais une overdose de citations. Et je ne trouve pas celle-ci particulièrement éclairante pour ton propos. Certes il y a un lien entre ce que tu dis et ce que dit cette citation, mais elle n'est pas vraiment une illustration de ce que tu dis. Du coup elle brouille le message}
Although being an important insight about the (likely vain) hope of determining a single correct recommendation in many complex decision situations (about which more below), this answer \commentYM{quelle answer ? à quelle question ? je suis perdu}does not address the problem that some principles of decision making may be wrong guides in some situations, possibly including some principles that are commonly applied in the field, hence the need of principled ways of discriminating them. \commentYM{est-ce vraiment ton projet ? as-tu l'intention de trouver la meilleure des axiomatiques de décision ? si c'est ton projet, il faut le dire clairement. Ce que je comprennais jusqu'ici était que tu cherchais un moyen de connaître les "preferences that represent informed, pondered decisions, that would be taken after due consideration by the individual", ce qui me semble être un tout autre projet}

\subsection{Some other related approaches}
\commentYM{related to what? to your approach or to the one explored above?}
The prescriptive approach to decision making \citep{bell_decision_1988} has been proposed (in supplement to the normative one and the descriptive one), which insists on incorporating the findings of psychologists about decision maker’s weaknesses of reasoning in the procedures leading to recommendations. I do not discuss this approach specifically here, because its methods can be considered to fall into the two categories I have described here above when it aims at obtaining reflective preference models \commentYM{j'ai rien compris. L'approche prescriptive tombe DANS LES DEUX CATEGORIES EN MEME TEMPS ? Il faut expliquer ce que tu racontes là}. When considering normative economics broadly, as I do here, as the methods aiming for valid prescription for an individual or a group, prescriptive approaches are a part of normative economics \commentYM{dans le papier sur la rationalité avec Alexis, on discute les faiblesse de la catégorisation de Bell et al., et je pense que ce qu'on dit va dans ton sens. Tu pourrais nous citer : ça me permettrait d'avoir 1 citation : personne ne cite mes papiers... snif... Plus sérieusement, je me demande ce que la normative ECONOMICS vient faire ici. Selon moi, le prescriptif sensu Bell et al. est une sous-catégorie du normatif sensu Bell et al.}.

Some authors have proposed to ground \commentYM{base} normative theories on observable aspects of nature and individuals, to solve some of the problems mentioned above; for example, \citet{kahneman_back_1997} propose to measure “experienced utility”, considered as a hedonic quality. \commentYM{il me semble qu'il faudrait être + précis ici : tous les problèmes mentionnées + haut ne peuvent pas être résolus par cette approche. Certains peuvent l'être, et encore, ça se discute, d'autres ne peuvent pas l'être. Si tu ne dissèques pas cela, ton affirmation semble creuse}To the best of my knowledge, these proposals never aim at observing reflective preferences as considered here and thus fall outside the scope of this article. \commentYM{tout à fait d'accord, mais du coup la manière dont tu les mentionnes les met trop en avant, surtout vu que tu dis qu'ils prétendent traiter certains des problèmes que tu explores précédemment. Il faudrait trouver une meilleure manière de faire référence à ces approches (on te reprocherait de ne pas les discuter), en passant, et en montrant plus clairement qu'elles sont hors champs ; probablement + en amont dans le texte}

The field of formal argumentation theory has developed technical approaches to deal with arguments. Such endeavor starts with an “attack” relation, which indicates which arguments argue against which. Various criteria can then be mathematically defined, in the form of a function selecting subsets of arguments given an attack relation, typically representing sets of arguments that “best resist” their counter-arguments. This field has greatly inspired this proposal, in particular, the notion of attack and of reinstatement. This field however does not view the attack relation as observable, but considers it as given, and its methodology is therefore not empirical in the sense seeked for here. (\Citet{cailloux_formal_2019} elaborates on this difference of objective.) \commentYM{ici aussi, je trouve la mention de ces approches maladroite, car elles sont mises en exergue comme si elles étaient des manières de traiter le problème qui t'intéresse, ce qu'elles ne sont pas}

An important literature exists about “debiasing techniques”: techniques that aim at encouraging individuals to think more about their decision problem in order to reduce occurrence of behavioral “mistakes”. Examples include suggesting the individuals to consider alternative points of view, or asking them to give reasons for their choices (an insightful article by \citet{mitchell_libertarian_2004} cite many such studies). This proposal \commentYM{de quoi tu parles ? de TON proposal ? pas clair} is situated at another level \commentYM{qu'est-ce que ça veut dire ?}, which could be considered as a way of justifying and comparing such debiasing techniques with no need of a prior agreement on what constitutes a behavioral mistake.\commentYM{très cryptique. Je n'ai rien compris à ce §, et je me demande pourquoi tu ne traites pas des debiasing techniques dans la même sous-section que les laundering techniques, qui me semblent poser les mêmes problèmes}

The current proposal pursues a path started with \citet{cailloux_formal_2019}, an article that defines the notion of \ac{DJ} and models that enable interactions with individuals whose \acp{DJ} are to be captured. That article however requires as fundamental observable a “trump” relation, indicating whether an individual thinks that an argument is against another argument; and a “support” relation, indicating whether an individual thinks that an argument supports a proposition.\commentYM{je ne pense pas qu'on puisse tenir pour acquis que le lectorat connait par coeur l'article delibenv. Il faut du coup donner plus d'explication. Par ailleurs, il est impératif que tu expliques pourquoi tu considères qu'il n'est pas raisonnable de considérer que trump et support sont observables} The current proposal defines every required concepts, such as the notion of arguments attacking other arguments, on the basis of a single observable that is compatible with the revealed preference approach priviledged in economic theory \commentYM{je ne pense pas qu'on puisse dire que l'approche RP est privilégiée en économie. Elle est aussi très critiquée. Si tu la considère particulièrement solide, il faut expliquer pourquoi}, in the sense that it can be defined by reference to what individuals effectively choose under various circumstances. The current proposal also defines explicitly (in the form of an axiom \commentYM{il n'y QU'UN axiome au final ?}) sufficient requirements to grant normative power to observations, an aspect which was previously left unclear. Finally, the current proposal links \ac{DJ} and deliberated preferences. \commentYM{indispensable de dire explicitement quelle est la différence entre les deux dès maintenant}

@@

With no intent of reducing the importance of these findings, it is important to qualify such statements. In supplement to them being not necessarily applicable to reflective preferences, the findings of psychologists do not, evidently, show that all preference statements are labile. As \citet[p.\ xvi]{lichtenstein_construction_2006} write, “To be sure, there are limits to the process of construction. Some televisions just won’t sell; most of the time, people are unlikely to want 80\% of their wages to go into savings. It would be an overstatement to say of preferences, as Gertrude Stein said of Oakland, that “there is no there there.” \commentYM{ce § et le précédent me semble mériter plus que le reste une réécriture. Par ailleurs, je ne vois pas à quoi ils servent dans cette sous-section}
